---
title: "Adjusting data + PCA"
author: "Carla Casanova"
date: "6/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

```{r libraries}
library(RadAR)
library(dplyr)
library(ggplot2)
library(factoextra)
library(datawizard)
```

## Preparing data for the analysis  

```{r load_rdr}
load("/Users/carla/Documents/Master Bioinformatics UAB/Pr치cticas Radiomics/Radiomic features/Results_rfeatures/results/radar_object_Norm_L1.rda")

rdr_L1
```

Filter radiomic features by original image type to adjust and perform the analysis only with this data:  

```{r filtering_image_type}
## filter by image types
rdr_filt_original <- filter_by_image_type(rdr = rdr_L1, image_type = c("original"))
 
rdr_filt_original
```

## Adjust by minimum pixel value  

Save the table with original values for radiomic features and prepare an object readable by `adjust()` function from *datawizard* package. Afterwards, adjust data by minimum pixel value variable:  

```{r adjust_all_features}
# Store table with all the features
table_radiomic_original <- assay(rdr_filt_original)

# Store col names and row names from the original table
patients_original <- colnames(table_radiomic_original) 
features_original <- rownames(table_radiomic_original)

# Let's prepare the data  
table_radiomic_original.T <- as.data.frame(t(table_radiomic_original))

# Change invalid column names
# Tags in original features
names(table_radiomic_original.T)[names(table_radiomic_original.T) == '10Percentile.original'] <- 'tenPercentile.original'
names(table_radiomic_original.T)[names(table_radiomic_original.T) == '90Percentile.original'] <- 'ninetyPercentile.original'

# Adjust all data for the variable image type
clinical_features_min_vox_adj_original <- adjust(table_radiomic_original.T, effect = "Minimum.original", keep_intercept = TRUE)

plot(clinical_features_min_vox_adj_original[,"Mean.original"], clinical_features_min_vox_adj_original[,"Variance.original"])
```

Before storing the adjusted data back to the `rdr` object, prepare the table for having the same format than the original data:  

```{r prepare_table_assay}
colnames(clinical_features_min_vox_adj_original) <- features_original
rownames(clinical_features_min_vox_adj_original) <- patients_original

assay_adjusted_original <- as.data.frame(t(clinical_features_min_vox_adj_original))
```

Finally, add a new `assay` to `rdr` object and save it:  

```{r add_assay_original}
# This code allows adding assays to rdr object
assays(rdr_filt_original)$adjusted_min <- assay_adjusted_original
```

To recover adjusted data, use the `assay()` function with the index number 2:  

```{r assay}
# Normal data
test <- assay(rdr_filt_original,1)

# Adjusted data
test2 <- assay(rdr_filt_original,2)
```

Save an `rdr` object with features from original images and with two assays (original and adjusted).  

```{r save_data_adjusted}
save(rdr_filt_original, file = "/Users/carla/Documents/Master Bioinformatics UAB/Pr치cticas Radiomics/Radiomic features/Results_rfeatures/radar_L1_Norm_adjusted_min.rda")
```

## Scaled PCA  

Loading data and recovering adjusted values by **minimum pixel value**:  

```{r table_adjusted_pca}
load("/Users/carla/Documents/Master Bioinformatics UAB/Pr치cticas Radiomics/Radiomic features/Results_rfeatures/radar_L1_Norm_adjusted_min.rda")

table_adjusted <- assay(rdr_filt_original, 2)

rdr_filt_original
```

When performing PCA we can see that most of the points are overlapped, meaning that only few groups of raidomic features are providing information:  

```{r pca_scaled}
prComp_scaled <- prcomp(table_adjusted, scale = TRUE)

# Plot two first principal components
plot(prComp_scaled$x[, 1], prComp_scaled$x[, 2], xlab = "PC1", ylab = "PC2", main = "Scaled PCA of radiomic features from original images")
```

Now, we'll see that PC1 + PC2 are explaining most of the variability:  

```{r check_PCA}
# Check effect of principal components 
summary(prComp_scaled)
```

```{r PCA_explained_variance_scaled}
fviz_eig(prComp_scaled)
```

This is an optional plot to see how feature types are distributed between the groups formed by principal components:  

```{r PCA_scaled_ggplot_feature_type}
# Store PCA data frame for features
PCA_scaled <- prComp_scaled$x

# Create a data frame with feature types
feature_names_PCA_scaled <- substr(rownames(PCA_scaled), 1, nchar(rownames(PCA_scaled))-9) 
# Check row data of rdr object which as feature name and feature type
ids <- rowData(rdr_filt_original)$feature_name %in% feature_names_PCA_scaled
out <- cbind(feature_names_PCA_scaled,
              rowData(rdr_filt_original)[ids, ]$feature_type)

# Add feature type to a data frame with PCA results
density.plot.PCA.scaled <- as.data.frame(PCA_scaled) %>%
  mutate(Feature_type = out[,2])

# Plot PC1 by feature type to find overlapped points
ggplot(data = density.plot.PCA.scaled) +
    geom_histogram(aes(x = PC1, fill = Feature_type)) +
  labs(y = "Frequency", title = "PC1 of scaled radiomic features from original images", x = "PC1", fill = "Feature type") +
  scale_fill_brewer(palette="PiYG")
```

Let's add principal components to `colData` of the `rdr` object. In this case, patients are rows in the `colData` table, so we need to add principal components for each patient, not per radiomic feature. These results are stored in the `rotation` section of the `prComp` object:  

```{r add_PCA_rdr}
# Save a data frame with PCA results for each patient (PCA by columns)
PCA_scaled_patients <- prComp_scaled$rotation

# Check patients are ordered in both data frames
identical(rownames(PCA_scaled_patients), rownames(colData(rdr_filt_original)))

# Add columns with PCAs
colData(rdr_filt_original) <- cbind(colData(rdr_filt_original), PCA_scaled_patients)
```

Finally, let's save the `rdr` object with PCA computed:  

```{r save_data_adjusted}
save(rdr_filt_original, file = "/Users/carla/Documents/Master Bioinformatics UAB/Pr치cticas Radiomics/Radiomic features/Results_rfeatures/radar_L1_Norm_adjusted_min_PCA.rda")
```

